On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-08-03 Manuscript/manuscript.tex
Fixed important issues with bib file and cleaned up manuscript

\subsection{Rank-based Comparison}\label{sec:rank}

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we use random permutation to obtain the p-value of the observed
test statistics $W^*$. This method is
available from the \texttt{clusWilcox.test()} function with
\texttt{method = `ds'} and \texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

@@

The rank based methods described in Section~\ref{sec:rank} are used
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  This is
a two group comparison performed two times: once comparing 2019 World
Championships to 2022 World Championships and can be found in 
Section~\ref{sec:results2019vs2022} and secondly comparing 2022 national
championships to the 2022 World Championships and can be found in
Section~\ref{sec:resultsnational}.

\jy{use two-sided}

\jy{include sample sizes; permutation number (B);}

\begin{table}
  \centering
  \caption{}
  \begin{tabular}{c c c c c} 
   \toprule
   Comparison & $exact = True$ & $exact \neq True$ & Sample size & Clusters/Athletes \\ 
   \midrule
   2019 vs 2022 & $2.9\cdot10^{-3}$ & $2.3\cdot10{-2}$ & 65 & 15 \\ 
   Nat vs Inter & $2.2\cdot10^{-16}$ & $3.1\cdot10^{-4}$ & 61 & 15 \\
   \bottomrule
  \end{tabular}
  \label{tab:Clusrankresults}
\end{table}

Table~\ref{tab:Clusrankresults} shows the results generated from four tests in R.
We use a \texttt{clusWilcox.test()} and set the permutation number (B) to be
1000 as the sample size in this analysis is small.  The sample size indicates
the total number of observations while clusters represnts the number of athletes
in each dataset.  It is worth noting that it is entirely coincidental that there
are fiftenn athletes in sets.

@@

One conclusion from the analysis in \citet{zhang2021correlation} was that 
athletes' reaction times improved from 2011 to 2019.  However this study does 
not consider data from 1999 to 2009, and over this timespan athletes get slower 
and then faster.  Figure~\ref{fig:Boxplots} shows that for the men's pooled data,
the year with the fastest median reaction time before 2022 was 1999.  Thus, this 
reject's Zhang et al.'s conclusion that athletes are getting faster because they 
were fast over twenty years ago. One possible explanation for why 2011 had such
high variation was because of the IAAF rule discussed earlier that increased the
penalty of a false start to an instant disqualification \citep{iaaf2009falsestart}.
Athletes may have been overly cautious to false start and may have consciously
or sub-consciously reacted slower as a result.  Instead, a more appropriate 
conclusion may be that there is a lot of variability in reaction times and that
the context of the races matters.

@@

Since 1985 Seiko Holding Corporations has served every World Athletics 
Championships as the official timer \citep{wa2022seiko}.  Reaction times were
not officially measured until 1999 in the men's 110 meter hurdles and women's
100 meter hurdles. Seiko's technology for detecting 
reaction times relies on the pressure that athletes exert on the plate when they 
push off.  Their systems measure the time differential between when the ``gun" 
goes off to start the race and when the pressure changes \citep{wa2022seiko}.  
If an athlete has a reaction time under 0.1 seconds, it is deemed a false start 
as it is considered that no athlete can react so quickly \citep{Seiko-Timing}.  
Thus, a reaction time of for example 0.05, suggests that the athlete predicted 
the gun and it was luck that caused their abnormally high time.  At the World 
Championships level, World Athletics want to remove that element of 
luck and thus impose the 0.1 second barrier.
\eds{I would argue that this paragraph should go in the discussion as there are 
no results presented here}
\of{I agree so I moved it here. I think it also could fit into the top of the
data section as in section~\ref{sec:Data}. Thoughts?}

\eds{also consider moving the following paragraph to discussion??}
As Seiko timed both events, there should not be a significant
difference if the timing equipment in both instances was working correctly.
\eds{you just mentioned in the last subsection that Seiko upgraded its 
technology prior to 2022}
of{Yes I agree}
To put it simply, there is not a reasonable explanation for 
this. It is inconceivable that so many athletes would improve such a significant
amount in such a small amount of time. Six countries are represented in this 
comparison, which drastically decreases the chances of this issue being caused by
one specific country.  The United States athletes' improvement in times was
discussed earlier in the paper, but this data shows that on average there was
\jy{any reference?}
improvement for British, Polish, French, Brazilian, and Spanish athletes.

@@

If there are issues with reaction time such as athletes approaching the 0.1 
second barrier again at the next World Championships (August 2023), then World 
Athletics needs to adjust that threshold and allow for faster reaction times so 
that athletes with superb reflexes are not penalized.  World Athletics did not
do anything in response to the \citet{komi2009iaaf} study, but perhaps they 
should now so that the story of Devon Allen is not repeated.

>>>>

\subsection{Rank-based Comparison}\label{sec:rank_methods}

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we use random permutation, which simulates the observed test 
statistic: $W^*$ given number of times to obtain the p-value of said test 
statistics. This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} and \texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

@@

The rank based methods described in are used in section~\ref{sec:rank_methods}
to determine if the reaction times for athletes who had competed at other 
championships differed significantly from the 2022 World Championships.  This is
a two group comparison performed two times: once comparing 2019 World
Championships to 2022 World Championships and then comparing 2022 national
championships to the 2022 World Championships.

\jy{use two-sided}
\of{Resolved}

\jy{include sample sizes; permutation number (B);}
\of{Resolved}

Table~\ref{tab:Clusrankresults} shows the results generated from four tests in R.
We use a \texttt{clusWilcox.test()} with \texttt{exact = TRUE} and 
\texttt{B = 1000} indicating that the random permutation size was chosen to be 
1000.  As the sample sizes are small: 65 and 49, we use \texttt{exact = TRUE} as
that is preferred.  In both cases, the national versus international comparison
produced stronger p values than the 2019 versus 2022 comparison, with the
national vs international \texttt{exact = TRUE} test producing a p value of 
$2.2\cdot10^{-16}$, which is the smallest possible value acheivable in R.  This
is a highly significant test result and shows that athletes who competed at the
2022 Track and Field World Championships were on average, significantly faster
at reacting than they were a few months prior.  For the 2019 versus 2022
same athlete comparison with \texttt{exact = TRUE} argument, the p value was
significant at $\alpha = 0.01$, however for when \texttt{exact = FALSE}, the p
value was higher and only significant at $\alpha = 0.05$. Nonetheless, these
tests help to show that athletes competing at the 2022 Track and Field World
Championships had faster reaction times than when they had competed previously.

\begin{table}
  \centering
  \caption{}
  \begin{tabular}{c c c c c} 
   \toprule
   Comparison & $exact = True$ & $exact \neq True$ & Sample size & Clusters/Athletes \\ 
   \midrule
   2019 vs 2022 & $2.9\cdot10^{-3}$ & $2.3\cdot10{-2}$ & 65 & 15 \\ 
   Nat vs Inter & $2.2\cdot10^{-16}$ & $1.0\cdot10^{-3}$ & 49 & 11 \\
   \bottomrule
  \end{tabular}
  \caption{2019 vs 2022 compares data from the same athletes who competed at the
  2019 and 2022 World Track and Field Championships. Nat vs Inter compares data
  from the same athletes who competed at the 2022 national-level championships
  and the 2022 World Track and Field Championships}
  \label{tab:Clusrankresults}
\end{table}

@@

%One conclusion from the analysis in \citet{zhang2021correlation} was that 
%athletes' reaction times improved from 2011 to 2019.  However this study does 
%not consider data from 1999 to 2009, and over this timespan athletes get slower 
%and then faster.  Figure~\ref{fig:Boxplots} shows that for the men's pooled data,
%the year with the fastest median reaction time before 2022 was 1999.  Thus, this 
%reject's Zhang et al.'s conclusion that athletes are getting faster because they 
%were fast over twenty years ago. One possible explanation for why 2011 had such
%high variation was because of the IAAF rule discussed earlier that increased the
%penalty of a false start to an instant disqualification \citep{iaaf2009falsestart}.
%Athletes may have been overly cautious to false start and may have consciously
%or sub-consciously reacted slower as a result.  Instead, a more appropriate 
%conclusion may be that there is a lot of variability in reaction times and that
%the context of the races matters.

@@

%Since 1985 Seiko Holding Corporations has served every World Athletics 
%Championships as the official timer \citep{wa2022seiko}.  Reaction times were
%not officially measured until 1999 in the men's 110 meter hurdles and women's
%100 meter hurdles. Seiko's technology for detecting 
%reaction times relies on the pressure that athletes exert on the plate when they 
%push off.  Their systems measure the time differential between when the ``gun" 
%goes off to start the race and when the pressure changes \citep{wa2022seiko}.  
%If an athlete has a reaction time under 0.1 seconds, it is deemed a false start 
%as it is considered that no athlete can react so quickly \citep{Seiko-Timing}.  
%Thus, a reaction time of for example 0.05, suggests that the athlete predicted 
%the gun and it was luck that caused their abnormally high time.  At the World 
%Championships level, World Athletics want to remove that element of 
%luck and thus impose the 0.1 second barrier.
%\eds{I would argue that this paragraph should go in the discussion as there are 
%no results presented here}
%\of{I agree so I moved it here. I think it also could fit into the top of the
%data section as in section~\ref{sec:Data}. Thoughts?}

@@

%If there are issues with reaction time such as athletes approaching the 0.1 
%second barrier again at the next World Championships (August 2023), then World 
%Athletics needs to adjust that threshold and allow for faster reaction times so 
%that athletes with superb reflexes are not penalized.  World Athletics did not
%do anything in response to the \citet{komi2009iaaf} study, but perhaps they 
%should now so that the story of Devon Allen is not repeated.
